---
title: "Parsing GEO Data"
author: "Keith Baggerly"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
  md_document: 
    toc: true
  html_document:
    toc: true
    number_sections: true
    code_folding: show
---

-   [Overview](#overview)
    -   [Introduction](#introduction)
    -   [Data and Methods](#data-and-methods)
    -   [Results](#results)
-   [Load Libraries](#load-libraries)
-   [Background on the GEO Data](#background-on-the-geo-data)
-   [Parsing the GEO Data](#parsing-the-geo-data)
    -   [Loading the GEO Data](#loading-the-geo-data)
    -   [Examining SOFT Files](#examining-soft-files)
    -   [General Notes](#general-notes)
    -   [A Helper Parsing Function](#a-helper-parsing-function)
        -   [Get Constants](#get-constants)
        -   [Get Sample Info](#get-sample-info)
        -   [Get Data Values](#get-data-values)
        -   [Return Results](#return-results)
        -   [Put the Function Together](#put-the-function-together)
    -   [Parsing SOFT Files](#parsing-soft-files)
    -   [Combining and Fixing Data](#combining-and-fixing-data)
        -   [Creating a Tibble of Sample
            Information](#creating-a-tibble-of-sample-information)
        -   [Creating a Tibble of Data
            Values](#creating-a-tibble-of-data-values)
    -   [Saving the Tibbles](#saving-the-tibbles)
-   [Initial Exploratory Data
    Analysis](#initial-exploratory-data-analysis)
    -   [Consistency of Probeset
        Ordering](#consistency-of-probeset-ordering)
    -   [First Few Values](#first-few-values)
    -   [Locations of Control (AFFX)
        Probes](#locations-of-control-affx-probes)
    -   [Sample ranges](#sample-ranges)
    -   [Sample Pairwise Correlation
        Matrix](#sample-pairwise-correlation-matrix)

# Overview

## Introduction

[Potti et al](https://www.nature.com/articles/nm1491) built
predictors of drug response using microarray profiles of
the NCI60 cell lines. They validated these predictors using
several publicly available microarray datasets. One of these
datasets, posted to the Gene Expression Omnibus (GEO), dealt
with women with breast cancer who had been treated with 
single agent docetaxel ([Chang et al](https://www.ncbi.nlm.nih.gov/pubmed/12907009), 2003). 
This GEO dataset was split into 
two subsets: GSE349, comprised of samples from women
deemed "resistant" to docetaxel, and GSE350, from women
who were "sensitive".

Here, we

* document our understanding of the GEO data more fully, 
* parse the data supplied into summary tibbles to make 
  later analyses easier, and
* perform some basic exploratory data analyses 

## Data and Methods

We use the data pulled from the web with `01_gather_raw_data.R`.

We draw documentation from the GEO pages for GSE349 and GSE350,
from the initial paper by Chang et al, and from communication
with the Chang et al authors.

We write a function to parse the SOFT-format data files
obtained for GSE349 and GSE350, and extract tibbles of 
both sample information and data values. Per the authors, 
we fix the Sensitive/Resistant label for one sample.
We combine tibbles across the two GSE subsets to produce 
one set of tibbles for the GEO dataset. We put
the sample information in "geo_sample_info",
with columns for GSE number, sensitive/resistant status, 
GSM number (sample id at GEO), and title (sample id used
in the Chang et al paper). We put the data values in 
"geo_data_values", with columns for sample id, probeset id,
value, and initial probeset row index. 

We visually examine the values reported for the first few
probesets. We examine and plot basic summary statistics
(e.g., sample quantiles). We plot pairwise sample 
correlations in a heatmap. 

## Results

We walk through assembly of the SOFT parsing function
in detail. Using this function, we extract tibbles of 
sample information for both GSE349 and GSE350, which 
we then combine.

We save the tibbles produced in "results/geo_data.RData".

We also do some quick exploration. The minimum sample values
have been thresholded to a common value of 5.89822, but nothing
else about the samples looks immediately remarkable. 

# Load Libraries

First, we load the libraries we'll need for this analysis.

```{r load_libs, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)

```

# Background on the GEO Data

Potti et al describe using the NCI60 data to predict response
for various patient cohorts for which corresponding microarray
data were publicly available. One of these cohorts was a group
of 24 women with breast cancer who were treated with single
agent docetaxel; the first report on this cohort was by 
[Chang et al](https://www.ncbi.nlm.nih.gov/pubmed/12907009) 
in the Lancet in 2003. Chang et al dichotomized the patients
into "Sensitive" and "Resistant" subcohorts of sizes 11 and
13, respectively. These patient tumors were interrogated with 
the same type of microarray (Affymetrix U-95Av2) as the NCI60
panel, so matching gene measurements will be easier for this 
dataset than for others where genes were interrogated using
different strands of cDNA. The gene expression data were posted
to the Gene Expression Omnibus (GEO) as GSE349 (the resistant
cohort) and GSE350 (the sensitive cohort). As 
[supplementary information](http://www.thelancet.com/pb-assets/Lancet/extras/01art11086webtable.pdf), the authors provide a table (in pdf) of the 
expression values by sample for each of the 92 genes they found 
to be particularly important for distinguishing sensitive from 
resistant.

Interestingly, GSE349 has data for 14 samples, not 13, and 
GSE350 has data for 10 samples, not 11. Personal communication with 
one of the Chang et al authors (Sue Hilsenbeck) confirmed that 
one of sensitive samples had been uploaded with the resistant 
group by mistake. 

> Sample \#377 is mislabeled in the GEO DB. It is listed there as resistant, but in reality it was sensitive.

# Parsing the GEO Data

## Loading the GEO Data

Now we load the SOFT data in as character vectors.

```{r load_gse_data}

gse349_soft <- readLines(here::here("data", "GSE349_family.soft.gz"))
gse350_soft <- readLines(here::here("data", "GSE350_family.soft.gz"))

```

## Examining SOFT Files

We don't need or want all of the data in the SOFT files, 
so we need to examine and reorganize what we do want.

## General Notes

Each SOFT file is simply a concatentation of text blocks.
Each block begins with a line beginning with "^". Let's 
take a look at the starting lines for GSE349.

```{r gse349_start_lines}

gse349_soft[which(substr(gse349_soft, 1, 1) == "^")]

```

The blocks for DATABASE, SERIES, and PLATFORM deal more 
with annotation than with assay values, so we'll pass on
these for now in order to focus on the gene expression 
values. Within each of the sample blocks, there are 
a few lines of annotation beginning with "!Sample_"
and a descriptor, followed by a sample data table.
The data table is preceded by a row of "!sample_table_begin"
followed by a header row and then a tab-delimited table 
of information, one row per value. Rows immediately 
before the sample table indicate both the number of 
data rows ("!Sample_data_row_count"), which is 12625 
for all of the arrays examined here, and the number of 
columns (one line per column, each consisting of a 
hash followed by the column name).

We'll organize some some of the sample information into
a tibble, and then create a tibble of the data values.
As we're creating the latter tibble, we'll add another 
column which wasn't in the SOFT table to indicate the 
row index of the probeset within the SOFT sample data 
table. This information shouldn't be relevant for 
analysis, since we should be matching on probeset (gene) id, 
but we might find it useful. 

## A Helper Parsing Function

Since we're going to be doing the same parsing more
than once, we write a helper function to take the 
character vector ("gse_soft") returned by readLines 
and return a list with the information we want. 
Specifically, we extract numeric constants, 
sample information, and data values. 

### Get Constants

The specific constants we want are the number of 
arrays in the dataset and the number of measurements
in each array.

```{r get_constants, eval=FALSE}

n_arrays <- 
  length(grep("^!Series_sample_id", gse_soft))

n_array_values <-
  gse_soft[grep("^!Platform_data_row_count", gse_soft)]
n_array_values <-
  as.numeric(substr(n_array_values,
                    nchar("!Platform_data_row_count = ") + 1,
                    nchar(n_array_values)))

```

### Get Sample Info

The sample information we want is the GSE number (SERIES),
the GSE title (Sensitive or Resistant), the GSM name
for each array, and the title (name the Chang group used)
for each array. We save this info in a tibble.

```{r get_sample_info, eval=FALSE}

gse_series <- gse_soft[grep("^\\^SERIES", gse_soft)]
gse_series <- 
  substr(gse_series, nchar("^SERIES = ") + 1, nchar(gse_series))

gse_title <- gse_soft[grep("^\\!Series_title", gse_soft)]
gse_title <- 
  substr(gse_title, nchar("!Series_title = ") + 1, 
         nchar(gse_title))

gsms <- gse_soft[grep("^\\^SAMPLE", gse_soft)]
gsms <- 
  substr(gsms, nchar("^SAMPLE = ") + 1, nchar(gsms))

gsm_titles <- 
  gse_soft[grep("^\\!Sample_title", gse_soft)]
gsm_titles <- 
  substr(gsm_titles, nchar("!Sample_title = ") + 1,
         nchar(gsm_titles))

sample_info <- 
  tibble(
    gse = gse_series,
    status = gse_title,
    gsm = gsms,
    title = gsm_titles)

```

### Get Data Values

The data values we want are the (probeset, expression value)
pairs for each array. We also record the row index of the 
probeset in the set of array values. We save this information
in a tibble.

```{r get_data_values, eval=FALSE}

data_start_rows <- 
  which(gse_soft == "!sample_table_begin") + 2
## use +2 to skip over header row

data_values <- 
  rep("", n_arrays * n_array_values)

start_row <- 1
end_row   <- n_array_values

for(i1 in 1:n_arrays){
  data_values[start_row:end_row] <- 
    gse_soft[data_start_rows[i1]:
               (data_start_rows[i1] + 
                  n_array_values - 1)]
  start_row <- start_row + n_array_values
  end_row   <- end_row   + n_array_values
}

data_values <- 
  matrix(unlist(strsplit(data_values, "\t")), 
         ncol = 2, byrow = TRUE)

data_values <- 
  tibble(
    sample_id = rep(sample_info$gsm, 
                    each = n_array_values),
    probeset_id = data_values[, 1],
    value = as.numeric(data_values[, 2]),
    row_index = rep(c(1:n_array_values), n_arrays))

```

### Return Results

Having pulled everything together, we
return the results in a list. 

```{r return_results, eval=FALSE}

all_data <- 
  list(n_arrays = n_arrays,
       n_array_values = n_array_values,
       sample_info = sample_info,
       data_values = data_values)

all_data

```

### Put the Function Together

Having assembled all of the pieces, we put them
together for use.

```
parse_gse_soft_file <- function(gse_soft){

  <<get_constants>>
  <<get_sample_info>>
  <<get_data_values>>
  <<return_results>>  
    
}
```

```{r build_all, echo=FALSE}

parse_gse_soft_file <- function(gse_soft){

  <<get_constants>>
  <<get_sample_info>>
  <<get_data_values>>
  <<return_results>>  
    
}

```

## Parsing SOFT Files

Now we parse the SOFT files.

```{r get_values}

gse349 <- parse_gse_soft_file(gse349_soft)
gse350 <- parse_gse_soft_file(gse350_soft)

```

As a sanity check, we look at the information 
returned.

```{r check_values}

gse349$sample_info
gse350$sample_info
gse349$data_values
gse350$data_values

```

These look reasonable. 

## Combining and Fixing Data

We'd actually like to work with the data from both
datasets, so we can combine the sample information and
data value tibbles from the derived from the two SOFT 
files. We also need to relabel sample 377 (GSM4913)
as Sensitive.

### Creating a Tibble of Sample Information

Now we assemble the sample information for the combined dataset.

```{r combine_sample_info}

geo_sample_info <- 
  bind_rows(gse349$sample_info, gse350$sample_info)
geo_sample_info$status[geo_sample_info$title == "377"] <-
  "Sensitive"

```

### Creating a Tibble of Data Values

Now we assemble the data values for the combined dataset. 

```{r create_data_value_tibble}

geo_data_values <- 
  bind_rows(gse349$data_values, gse350$data_values)

```

## Saving the Tibbles

We save these tibbles in an RData file for later use.

```{r save_data}

save(geo_sample_info,
     geo_data_values,
     file = here::here("results", "geo_data.RData"))

```

# Initial Exploratory Data Analysis

As a sanity check, we do some quick skims of the data.

## Consistency of Probeset Ordering

We recorded both probeset id and row index for each of the
data values, but in general we would hope the data's being
presented in the same order across samples. Let's check 
if the probesets are in the same rows across samples. 

```{r check_probeset_ordering}

test_matches <- 
  geo_data_values %>% 
  select(probeset_id, row_index) %>% 
  group_by(probeset_id) %>% 
  summarise(all_match = all(row_index == row_index[1])) %>% 
  select(all_match) %>% 
  summarise(n_good_probes = sum(all_match)) 

unlist(test_matches)

```

The row indices are the same across all samples for
all 12625 probesets.

## First Few Values

We eyeball the values of the first 4 probesets across
samples to see if the samples look comparable. 

```{r first_few_values}

geo_first_data <-
  geo_data_values %>% 
  filter(row_index < 5)

geo_first_probeset_plots <- 
  ggplot(geo_first_data, 
         aes(sample_id, value, group = probeset_id,
             colour = probeset_id)) +
  geom_point() + 
  geom_line() +
  ylab("Gene Expression") + 
  xlab("") +
  ggtitle("GEO Expression Values for First 4 Probesets") +
  theme(axis.text.x = 
          element_text(angle = 90, vjust = 0.5, hjust = 1))

geo_first_probeset_plots

```

These look fine - no clear evidence of something like
one probeset having high values for one subset of samples
and low values for the others. The "MurlL2" probeset sure
looks like a negative control (values are consistently low).

## Locations of Control (AFFX) Probes

The first 4 probesets all have names beginning with "AFFX"
(Affymetrix), indicating these are "control" probes. There
should be 67 control probesets on the Affymetrix U95 arrays
used here, so these are probably in the first 67 positions.
Let's check this. 

```{r tabulate_affx_locations}

control_probes <- 
  geo_data_values %>% 
  filter(grepl("^AFFX", probeset_id))
table(control_probes$row_index)

```

Yep, all of the control probes are at the beginning. 

## Sample ranges

We also check some standard summary values (min, med, max)
in case something odd shows up. 

```{r get_ranges}

geo_sample_ranges <- 
  geo_data_values %>% 
  group_by(sample_id) %>% 
  summarise(q000_min = min(value),
            q050_med = median(value),
            q100_max = max(value))

geo_sample_ranges

geo_sample_ranges$q000_min

```

The minimum values are all tied at 5.89822, indicating
some type of thresholding effect is present. If there are
any issues with the other summaries, these aren't as 
readily apparent.

## Sample Pairwise Correlation Matrix

As a sanity check, we also examine the pairwise sample
correlations. Since all of these samples are of breast
tissue, we would expect most of the genes to have similar
expression levels, so correlations should be clearly positive. 

We do this by first mapping from the full data tibble to a 
gene by sample matrix of values, compute the correlation
matrix using the matrix implementation in `cor`, and then
map back to a tibble for plotting purposes. 

Let's compute the correlations.

```{r assemble_correlation_tibble}

geo_data_matrix <- 
  geo_data_values %>% 
  select(-starts_with("row")) %>% 
  spread(sample_id, value) %>% 
  select(-probeset_id) %>% 
  as.matrix()

geo_cor_matrix <- 
  cor(geo_data_matrix, geo_data_matrix, method = "spearman")

geo_cor_tibble <- 
  tibble(gsm_1 = rep(colnames(geo_cor_matrix), 
                     each = nrow(geo_cor_matrix)),
         gsm_2 = rep(rownames(geo_cor_matrix),
                     times = ncol(geo_cor_matrix)),
         corr = as.vector(geo_cor_matrix))

rm(geo_data_matrix, geo_cor_matrix)

```

Now we plot a correlation heatmap.

```{r plot_corr_heatmap}

geo_cor_heatmap <- 
  ggplot(geo_cor_tibble, aes(gsm_1, gsm_2)) + 
  geom_tile(aes(fill = corr), color = "white") + 
  scale_fill_viridis_c() + 
  ylab("") + 
  xlab("") + 
  ggtitle("GEO Pairwise Sample Rank Correlations") +
  theme(axis.text.x = 
          element_text(angle = 90, hjust = 1, vjust = 0.5))

geo_cor_heatmap  

```

In general, we see the positive correlations we expect. 
There are a few samples (GSM4905, GSM4909) which don't
line up with the others as well, but even here the 
correlations are above 0.5.