---
title: "Identifying the Potti Data Sources"
author: "Keith Baggerly"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
  md_document: 
    toc: true
  html_document:
    toc: true
    number_sections: true
    code_folding: show
---

-   [Overview](#overview)
    -   [Introduction](#introduction)
    -   [Data and Methods](#data-and-methods)
    -   [Results](#results)
    -   [Conclusions](#conclusions)
-   [Libraries](#libraries)
-   [Loading Data](#loading-data)
-   [Matching Potti to NCI60](#matching-potti-to-nci60)
    -   [Notes on the First Probeset](#notes-on-the-first-probeset)
    -   [Rearranging the Datasets to Matrix
        Form](#rearranging-the-datasets-to-matrix-form)
    -   [Counting Exact Matches When Rows Are Matched on Probeset
        ID](#counting-exact-matches-when-rows-are-matched-on-probeset-id)
    -   [Organizing the Exact Matches in a
        Tibble](#organizing-the-exact-matches-in-a-tibble)
    -   [Adding Sample Information, Identifying
        Replicates](#adding-sample-information-identifying-replicates)
-   [Matching Potti to GEO](#matching-potti-to-geo)
    -   [Matching the First Probeset](#matching-the-first-probeset)
    -   [Checking What Samples Are
        Matched](#checking-what-samples-are-matched)
    -   [Spot Checking the Next Few
        Probesets](#spot-checking-the-next-few-probesets)
    -   [Testing for Agreement
        Throughout](#testing-for-agreement-throughout)
    -   [Testing Agreement, Tweaking the Last Few
        Indices](#testing-agreement-tweaking-the-last-few-indices)
    -   [Bundle the Perfect Matches Into A
        Tibble](#bundle-the-perfect-matches-into-a-tibble)
-   [Combine Matches Across Datasets](#combine-matches-across-datasets)
-   [Save the Sources](#save-the-sources)


# Overview

## Introduction

Given the tibbles for the Potti, NCI60, and GEO datasets
assembled in analyses 02, 03, and 04, respectively, we'd
like to see if we can more precisely identify the sources
of each of the columns in the Potti table. 

The Potti predictors were nominally assembled from the NCI60
data, so we would expect the Potti columns to be some simple
function of some of the NCI60 columns. The values for all of 
the Cytoxan and Docetaxel samples are so different that we
suspect something else happened here. The minimum values 
for all of these samples are 5.89822. The minimum values
for all of the samples in the GEO dataset are also 5.89822,
so we suspect a link.

## Data and Methods

We load the previously assembled tibbles from the 
associated RData files. 

The first probeset reported in both the Potti and 
NCI60 tibbles is `36460_at`, and the values are 
given to 5 or 6 decimal places. Given this level of
precision, we first check for exact matches, and
whether such matches persist beyond the first probeset. 

The probeset ordering in the GEO data is different
than the Potti ordering, so instead we search for
exact matches to the first Potti probeset values in 
the Cytoxan and Docetaxel columns across the entire
GEO dataset. We repeat this process for the second
probeset, and check whether the joint results suggest
an exploitable pattern. 

We bundle information about the matches identified
into a tibble with rows for the Potti sample information
and the NCI60 and/or GEO sample information as 
appropriate. 

## Results

Matching rows on probeset id, we find perfect
matches for all Potti samples save those for Cytoxan and 
Docetaxel. All NCI60 samples identified are from the "A" 
set of replicates - given multiple replicate arrays, 
only the first was used. 

Using the first probeset values in the Potti Cytoxan and Docetaxel
columns, we find perfect matches for all of them in values for 
the 68th probeset reported for the GEO data; the first 67 probeset
rows correspond to control probes. Dropping the first 67 probesets
from the GEO dataset and checking agreement matching on row order, 
we get perfect matches for the first 12533 of 12558 probesets. 
The last 25 values also match after permuting the last few row
indices. Since matching was on row index and not
probeset id, the GEO dataset values were effectively randomly
scrambled relative to all the others.

We save a tibble with the mappings identified to 
"results/potti_data_sources.RData".

## Conclusions

We know where all of the Potti data came from.
The fact that the docetaxel validation data was 
supplied as the nominal source of both the cytoxan
and docetaxel predictors is disturbing. The probeset
order scrambling further renders the cytoxan and
docetaxel columns unusable. 

# Libraries

Here we load the libraries we'll need for this analysis.

```{r load_libs, message=FALSE, warning=FALSE}

library(tidyverse)
library(here)

```

# Loading Data

Next, we load the cleaned data we previously assembled
from the Potti, NCI60, and GEO datasets. 

```{r load_data}

load(here::here("results", "potti_data.RData"))
load(here::here("results", "nci60_data.RData"))
load(here::here("results", "geo_data.RData"))

```

# Matching Potti to NCI60

Now we try matching the Potti and NCI60 data columns.

## Notes on the First Probeset

The first probeset for which data was reported in 
both the Potti and NCI60 datasets was 36460\_at, 
suggesting the latter was likely the source for 
at least some of the former. Since the expression
values are reported to several decimal places, 
we may be able to identify the source samples 
based on exact matching. 

## Rearranging the Datasets to Matrix Form

First, we spread the data out into
a more matrix-like arrangement.

```{r spread_out_data}

n_potti_samples <- 134
n_nci60_samples <- 180

potti_data_matrix <-
  potti_data_values %>% 
  spread(key = sample_id, value = value) %>%
  arrange(row_index)

nci60_data_matrix <- 
  nci60_data_values %>% 
  select(sample_id, probeset_id, value) %>% 
  spread(key = sample_id, value = value)
nci60_data_matrix <- 
  nci60_data_matrix[match(potti_data_matrix$probeset_id,
                          nci60_data_matrix$probeset_id), ]

```

## Counting Exact Matches When Rows Are Matched on Probeset ID

Now, having lined the samples up by probeset id, we count
the number of exact matches between Potti samples and 
NCI60 samples. 

```{r match_on_first_probeset}

match_count <- 
  matrix(0, nrow = n_potti_samples, ncol = n_nci60_samples)
rownames(match_count) <- 
  names(potti_data_matrix)[3:(n_potti_samples + 2)]
colnames(match_count) <- 
  names(nci60_data_matrix)[2:(n_nci60_samples + 1)]

for(i1 in 1:n_potti_samples){
  for(i2 in 1:n_nci60_samples){
    match_count[i1, i2] <- 
      sum(potti_data_matrix[, i1 + 2] ==
            nci60_data_matrix[, i2 + 1])
  }
}

table(match_count)

```

We find perfect matches (all 12558 probeset values agree)
for 94 of the 134 Potti samples in the NCI60 data. 

## Organizing the Exact Matches in a Tibble

Let's look at these matches in more detail.
First, we arrange the matches in a tibble. 

```{r arrange_matches}

perfect_matches <- 
  which(match_count > 12000, arr.ind = TRUE)
perfect_match_nci60 <- 
  rep(NA, n_potti_samples)
perfect_match_nci60[perfect_matches[, "row"]] <- 
  colnames(match_count)[perfect_matches[, "col"]]

potti_nci60_tibble <- 
  tibble(potti_sample = rownames(match_count),
         nci60_sample = perfect_match_nci60)

```

## Adding Sample Information, Identifying Replicates

Next, we bundle all the information about 
which NCI60 samples are being used
by Potti et al for which drugs, and 
summarize the mappings.

```{r matches_by_drug}

potti_nci60_tibble <- 
  left_join(potti_nci60_tibble, potti_sample_info, 
            by = c("potti_sample" = "sample_id"))

potti_nci60_tibble <- 
  left_join(potti_nci60_tibble, nci60_sample_info,
            by = c("nci60_sample" = "sample_id"))

table(potti_nci60_tibble$rep_id,
      potti_nci60_tibble$drug_name,
      potti_nci60_tibble$contrast_group,
      useNA = "ifany")

```

All matched samples are from the "A" replicate set from the
NCI60, so given the triplicate reps, they simply chose the
first. We find matches for all of the samples used for 5 of 
the 7 drugs. We find no matches for any of the Cytoxan or
Docetaxel samples. This fits with the fact that exploratory 
data analysis of the Potti data matrix showed the samples
for these two drugs didn't fit with the other samples at all. 

# Matching Potti to GEO

When we looked at the individual datasets, we saw that the
minimum value for all of the GEO samples was 5.89822. The
minimum value for all of the Potti Cytoxan and Docetaxel
samples is also 5.89822, which suggests these Potti samples
are coming from the GEO dataset. 

## Matching the First Probeset

As with the NCI60 data, we can check for exact matches, 
but we can't assume the probeset ids will match, and we 
should try to avoid instances of the minimum value
for which there will be a lot of matches. Since we 
can't trust the probesets to match, we'll preserve the
original presentation ordering to the extent possible. 

```{r check_geo_matches}

geo_data_matrix <-
  geo_data_values %>% 
  spread(key = sample_id, value = value) %>%
  arrange(row_index)
n_geo_samples <- nrow(geo_sample_info)

n_potti_unmatched <- 
  sum(is.na(potti_nci60_tibble$nci60_sample))

cytox_sample_offset <- 
  grep("^Cytox", names(potti_data_matrix))[1] - 1

match_probe_1 <- 
  matrix(NA, nrow = n_potti_unmatched, ncol = n_geo_samples)
for(i1 in 1:n_potti_unmatched){
  for(i2 in 1:n_geo_samples){
    temp <- which(geo_data_matrix[, i2 + 1] == 
                    potti_data_matrix[[i1 + 
                      cytox_sample_offset]][1])
    if(length(temp) >= 1){
      match_probe_1[i1, i2] <- temp[1]
    }
  }
}

table(match_probe_1)

```

There are exact matches of probeset value 1
for all of the 40 unmatched Potti samples to probeset value
68 in the GEO data. An offset of 67 is suggestive,
because there are exactly 67 "control" probesets on 
the Affymetrix arrays, and these are in the first 67
positions in the original probeset ordering in the 
GEO data. Since the control probesets were stripped
from the Potti data, it may be that they simply skipped
the first 67 rows. 

## Checking What Samples Are Matched

Let's take a look at the first 
values in the Cytoxan samples and the 68th values in 
GEO as a sanity check (we'll check the Docetaxel 
values too for completeness). 

```{r check_first_values}

unlist(potti_data_matrix[1, grep("^Cytox", 
                                 names(potti_data_matrix))])

unlist(geo_data_matrix[68, 3:26])

match(unlist(potti_data_matrix[1, grep("^Cytox", 
                                 names(potti_data_matrix))]),
      unlist(geo_data_matrix[68, 3:26]))

match(unlist(potti_data_matrix[1, grep("^Doce", 
                                 names(potti_data_matrix))]),
      unlist(geo_data_matrix[68, 3:26]))

```

Interestingly, not only do the values match, but the
sequences of values are sequentially aligned in two 
subsequences of length 10. Since the 0 and 1 contrast 
groups both have 10 members, this indicates the cell
lines were chosen sequentially from the GEO dataset
within contrast groups. Since the Docetaxel and Cytoxan
groups are the same modulo contrast group labelings, 
that's where these samples come from there too. 

The GEO dataset does involve the patient responses 
to treatment with single-agent docetaxel, so there
is some potential linkage for this drug, but the GEO
dataset was supposed to be used as the test set here, 
not the validation set. There's no rationale for 
cytoxan at all. 

## Spot Checking the Next Few Probesets

Let's take a look at the next few probesets. 

```{r check_first_few_cytox}

potti_data_matrix[1:5, c("probeset_id", "Cytox_0_075")]

geo_data_matrix[c(1:5) + 67, c("probeset_id", "GSM4901")]

```

The values continue to track - after skipping the control
probesets, they took the values as given. But, because the
probeset order was different, the values were effectively
scrambled relative to what they should have been. 

## Testing for Agreement Throughout

Does this continue throughout?

```{r }

sum(potti_data_matrix[, "Cytox_0_075"] ==
      geo_data_matrix[c(1:12558) + 67, "GSM4901"])

which(potti_data_matrix[, "Cytox_0_075"] !=
      geo_data_matrix[c(1:12558) + 67, "GSM4901"])

match(unlist(potti_data_matrix[12534:12558, "Cytox_0_075"]),
      unlist(geo_data_matrix[c(12534:12558) + 67, "GSM4901"]))

```

The first 12533 values match in the order presented. The
last 25 values in the Potti sample also match the last 25
values in the GEO sample, but the order has been scrambled. 
If this was just a one-off scrambling though, fixing things
here should also fix things for other matching pairs. 

## Testing Agreement, Tweaking the Last Few Indices

Let's check whether fixing the ordering of the last few
probesets in one sample fixes the ordering for all samples.

```{r test_potti_geo_order}

potti_geo_order <- 
  c(1:12533,
    match(
      unlist(geo_data_matrix[c(12534:12558) + 67, "GSM4901"]),
      unlist(potti_data_matrix[12534:12558, "Cytox_0_075"])) +
      12533)

sum(potti_data_matrix[potti_geo_order, "Cytox_0_075"] ==
      geo_data_matrix[c(1:12558) + 67, "GSM4901"])

sum(potti_data_matrix[potti_geo_order, "Cytox_0_076"] ==
      geo_data_matrix[c(1:12558) + 67, "GSM4902"])

```

Yep, after this one minor scramble at the end, the Cytoxan
values appear to match the GEO values across the board. 
Let's confirm this level of agreement holds throughout. 

```{r match_potti_to_geo}

match_geo_probes <- 
  matrix(NA, nrow = n_potti_unmatched, ncol = n_geo_samples)

for(i1 in 1:n_potti_unmatched){
  for(i2 in 1:n_geo_samples){
    match_geo_probes[i1, i2] <- 
      sum(potti_data_matrix[potti_geo_order, 
                            i1 + cytox_sample_offset] ==
            geo_data_matrix[c(1:12558) + 67, i2 + 2])
  }
}

table(match_geo_probes[match_geo_probes > 1000])
table(apply(match_geo_probes, 1, max))

```

There are 40 perfect matches of Potti columns to GEO columns,
one for each of the 40 Potti columns that didn't match any
NCI60 columns. 

## Bundle the Perfect Matches Into A Tibble

Let's extract the matches.

```{r arrange_geo_matches}

perfect_matches_geo <- 
  which(match_geo_probes > 12000, arr.ind = TRUE)
perfect_match_geo <- 
  rep(NA, n_potti_unmatched)
perfect_match_geo[perfect_matches_geo[, "row"]] <- 
  names(geo_data_matrix)[3:26][perfect_matches_geo[, "col"]]

potti_geo_tibble <- 
  tibble(potti_sample = 
           names(potti_data_matrix)[c(1:n_potti_unmatched) +
                                      cytox_sample_offset],
         geo_sample = perfect_match_geo)

potti_geo_tibble

```

# Combine Matches Across Datasets

Now lets combine this information with the earlier tibble
mapping Potti samples to NCI60 samples. 

```{r combine_nci60_and_geo}

potti_geo_tibble <- 
  left_join(potti_geo_tibble,
            geo_sample_info,
            by = c("geo_sample" = "gsm"))

potti_data_source_tibble <- 
  left_join(potti_nci60_tibble,
            potti_geo_tibble,
            by = "potti_sample")

```

# Save the Sources

We save the mapping tibble to an RData file for
later use if needed. 

```{r save_mappings}

save(potti_data_source_tibble,
     file = here::here("results", 
                       "potti_data_sources.RData"))

```

